---
title: 'XAI – Exercise 5: Partial Dependence Plots (PDP)'
author: "Data Science – Model-Agnostic Methods"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 2
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width  = 10,
  fig.height = 5
)
```

---

# Introduction

This report applies **Partial Dependence Plots (PDP)** to two real-world prediction tasks. PDP is a model-agnostic global interpretability method that visualises the marginal effect of one or two features on the predicted outcome of a previously fitted model. All models are **Random Forests** fitted using the `randomForest` package.

The partial dependence function is estimated as:

$$\hat{f}_{x_S}(x_S) \approx \frac{1}{n}\sum_{i=1}^{n}\hat{f}(x_S,\, x_C^{(i)})$$

where $x_S$ is the feature of interest and $x_C$ are the remaining features marginalised over the data. For every plot, a grid of **at least 50 evenly spaced values** is evaluated. Rug marks along the axes represent the empirical distribution of training observations, helping distinguish well-supported regions from potential extrapolations.

---

# Exercise 1 – 1D PDP: Bike Rental Prediction

## Data Preparation and Model Training

```{r ex1-setup}
# Libraries
library(randomForest)
library(ggplot2)
library(gridExtra)
library(scales)

# Load data
day <- read.csv("day.csv")

# Create days_since_2011 from the date column
day$days_since_2011 <- as.numeric(
  as.Date(day$dteday) - as.Date("2011-01-01")
)

# Feature set (exclude raw date, id, casual, registered)
features_bike <- c("days_since_2011", "season", "yr", "mnth", "holiday",
                   "weekday", "workingday", "weathersit",
                   "temp", "atemp", "hum", "windspeed")

train_bike <- day[, c(features_bike, "cnt")]

# Fit Random Forest on full bike dataset (731 rows — manageable)
set.seed(42)
rf_bike <- randomForest(cnt ~ ., data = train_bike, ntree = 500)

cat("Random Forest – Bike Rentals\n")
cat("OOB R²:", round(1 - rf_bike$mse[500] / var(day$cnt), 3), "\n")
```

## PDP Helper Functions

```{r ex1-helpers}
# Generic 1D PDP: evaluates model predictions over a grid of >= 50 values
compute_pdp_1d <- function(model, feature, train_df, n_grid = 50) {
  grid_vals <- seq(min(train_df[[feature]]),
                   max(train_df[[feature]]),
                   length.out = n_grid)
  pdp_vals <- sapply(grid_vals, function(v) {
    tmp            <- train_df
    tmp[[feature]] <- v
    mean(predict(model, newdata = tmp))
  })
  data.frame(x = grid_vals, yhat = pdp_vals)
}

# Plot function matching the class slide style (line + rug)
plot_pdp_bike <- function(pdp_df, x_label, train_df, feature,
                          y_limits = c(0, 5500)) {
  ggplot(pdp_df, aes(x = x, y = yhat)) +
    geom_line(colour = "black", linewidth = 0.9) +
    geom_rug(
      data        = train_df,
      aes_string(x = feature),
      sides       = "b",
      alpha       = 0.4,
      colour      = "black",
      inherit.aes = FALSE
    ) +
    scale_y_continuous(limits = y_limits,
                       breaks = seq(0, 5000, by = 1000)) +
    labs(x = x_label, y = "Predicted number of bike rentals") +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}
```

## Results

```{r ex1-plots, fig.height=8, fig.width=10}
# Compute the four 1D PDPs (50 grid points each)
pdp_days <- compute_pdp_1d(rf_bike, "days_since_2011", train_bike)
pdp_temp <- compute_pdp_1d(rf_bike, "temp",            train_bike)
pdp_hum  <- compute_pdp_1d(rf_bike, "hum",             train_bike)
pdp_wind <- compute_pdp_1d(rf_bike, "windspeed",       train_bike)

# Build plots
p1 <- plot_pdp_bike(pdp_days, "Days since 2011", day, "days_since_2011")
p2 <- plot_pdp_bike(pdp_temp, "Temperature",     day, "temp")
p3 <- plot_pdp_bike(pdp_hum,  "Humidity",        day, "hum")
p4 <- plot_pdp_bike(pdp_wind, "Windspeed",       day, "windspeed")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Analysis

**Days since 2011** shows a clear positive and roughly monotone trend. Predicted rentals grow steadily over time, reflecting the growing adoption of the service throughout 2011 and 2012. The Random Forest has captured a strong temporal effect, likely encoding both seasonal patterns and year-on-year demand growth. The rug marks confirm that observations span the entire time range without gaps.

**Temperature** exhibits a non-linear, inverted-U relationship. Rentals increase steeply from cold values toward moderate temperatures, reach a peak, and then level off or slightly decline at the highest temperatures. Both extreme cold and excessive heat discourage cycling, and the model captures this behaviour effectively.

**Humidity** has a predominantly negative effect. Predicted rentals remain relatively stable at low-to-moderate humidity levels but drop noticeably as humidity increases, consistent with rainy or uncomfortable atmospheric conditions deterring users. The rug marks show that most observations cluster in the mid-range, where model estimates are best-supported.

**Windspeed** displays a clear and sustained negative relationship: as wind increases, predicted rentals decline consistently. Strong winds make cycling physically more demanding and less pleasant, a behavioural pattern the model has learned across the full range of observed wind values.

---

# Exercise 2 – 2D PDP: Humidity × Temperature

## Approach

The 2D PDP evaluates predictions on a **50 × 50 grid** (2 500 combinations) of joint humidity and temperature values. The marginal expectation is approximated using a random sample of **500 observations** to keep computation efficient. `geom_tile()` is used to render the heatmap, with tile dimensions slightly expanded (`× 1.05`) to avoid visual gaps. Rug marks on both axes show where real data actually concentrates.

```{r ex2-pdp, fig.height=6, fig.width=8}
# Draw 500-row sample for PDP marginalisation
set.seed(42)
sample_idx_2d <- sample(nrow(train_bike), 500)
train_2d      <- train_bike[sample_idx_2d, ]

# Build 50 × 50 grid
hum_grid  <- seq(min(train_bike$hum),  max(train_bike$hum),  length.out = 50)
temp_grid <- seq(min(train_bike$temp), max(train_bike$temp), length.out = 50)
grid_2d   <- expand.grid(hum = hum_grid, temp = temp_grid)

# For each grid cell, substitute hum and temp and average predictions
grid_2d$yhat <- apply(grid_2d, 1, function(row) {
  tmp      <- train_2d
  tmp$hum  <- row["hum"]
  tmp$temp <- row["temp"]
  mean(predict(rf_bike, newdata = tmp))
})

# Step sizes to avoid holes in geom_tile
hum_step  <- (max(hum_grid)  - min(hum_grid))  / (length(hum_grid)  - 1)
temp_step <- (max(temp_grid) - min(temp_grid)) / (length(temp_grid) - 1)

# 2D PDP heatmap with marginal rug distributions
ggplot(grid_2d, aes(x = hum, y = temp, fill = yhat)) +
  geom_tile(width  = hum_step  * 1.05,
            height = temp_step * 1.05) +
  scale_fill_viridis_c(
    name   = "Predicted\nbike rentals",
    option = "viridis"
  ) +
  geom_rug(
    data        = day,
    aes(x = hum, y = temp),
    sides       = "bl",
    alpha       = 0.35,
    colour      = "white",
    inherit.aes = FALSE
  ) +
  labs(
    title = "2D Partial Dependence Plot – Humidity × Temperature",
    x     = "Humidity",
    y     = "Temperature"
  ) +
  theme_bw(base_size = 12) +
  theme(panel.grid = element_blank())
```

## Interpretation

The 2D PDP reveals the **joint marginal effect** of humidity and temperature on predicted bike rentals:

- **High temperature combined with low humidity** (bright/yellow region) produces the highest predicted demand — these are the most favourable cycling conditions.
- **Low temperature combined with high humidity** (dark/purple region) yields the lowest predictions, as cold and wet conditions strongly discourage use.
- The colour gradient follows a clear diagonal from bottom-right (worst) to top-left (best), confirming that both features act in a **reinforcing manner**: high humidity amplifies the negative effect of cold temperatures, and vice versa.
- The white rug marks show that most real observations cluster in the mid-ranges of both axes. Predictions in the extreme corners (very dry + very warm, or very wet + very cold) rely on fewer supporting data points and should be interpreted with caution.

---

# Exercise 3 – PDP: House Price Prediction

## Data Preparation

The King County house dataset contains 21 613 rows. To keep computation tractable, a **single random subset of 2 000 rows** is drawn from the full dataset. This same subset is used both to train the Random Forest and to compute the PDPs.

```{r ex3-setup}
# Load data
houses <- read.csv("kc_house_data.csv")

# Ensure numeric types
houses$floors    <- as.numeric(as.character(houses$floors))
houses$bathrooms <- as.numeric(houses$bathrooms)

# Feature set as specified in the exercise
features_house <- c("bedrooms", "bathrooms", "sqft_living",
                    "sqft_lot", "floors", "yr_built")

# Random subset of 2 000 rows — used for both training and PDP computation
set.seed(42)
idx_house   <- sample(nrow(houses), 2000)
train_house <- houses[idx_house, c(features_house, "price")]

# Fit Random Forest on the subset
set.seed(42)
rf_house <- randomForest(price ~ ., data = train_house, ntree = 500)

cat("Random Forest – House Prices\n")
cat("OOB R²:", round(1 - rf_house$mse[500] / var(train_house$price), 3), "\n")
```

## PDP Computation and Plots

```{r ex3-pdp, fig.height=8, fig.width=10}
# Plot function for house prices
plot_pdp_house <- function(pdp_df, x_label, ref_df, feature) {
  ggplot(pdp_df, aes(x = x, y = yhat)) +
    geom_line(colour = "black", linewidth = 0.9) +
    geom_rug(
      data        = ref_df,
      aes_string(x = feature),
      sides       = "b",
      alpha       = 0.4,
      colour      = "black",
      inherit.aes = FALSE
    ) +
    labs(x = x_label, y = "Predicted house price (USD)") +
    scale_y_continuous(labels = comma) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}

# Compute 1D PDPs using the same subset (50 grid points each)
pdp_bed   <- compute_pdp_1d(rf_house, "bedrooms",    train_house)
pdp_bath  <- compute_pdp_1d(rf_house, "bathrooms",   train_house)
pdp_sqft  <- compute_pdp_1d(rf_house, "sqft_living", train_house)
pdp_floor <- compute_pdp_1d(rf_house, "floors",      train_house)

# Build plots
ph1 <- plot_pdp_house(pdp_bed,   "Bedrooms",    train_house, "bedrooms")
ph2 <- plot_pdp_house(pdp_bath,  "Bathrooms",   train_house, "bathrooms")
ph3 <- plot_pdp_house(pdp_sqft,  "Sqft Living", train_house, "sqft_living")
ph4 <- plot_pdp_house(pdp_floor, "Floors",      train_house, "floors")

grid.arrange(ph1, ph2, ph3, ph4, ncol = 2)
```

## Analysis

**Bedrooms** shows a non-linear relationship. Predicted prices tend to increase from 1–3 bedrooms, but the effect flattens or even reverses at higher counts. A very high number of bedrooms in a similarly sized home may imply smaller individual rooms and lower quality, reducing marginal value. The rug marks confirm that most homes have between 2 and 5 bedrooms, making predictions at the extremes less reliable.

**Bathrooms** exhibits a broadly positive trend: more bathrooms are associated with higher predicted prices. Bathrooms serve as a quality proxy — homes with more bathrooms tend to be larger and more upscale. The effect is steepest at lower counts and gradually flattens beyond 3–4 bathrooms, suggesting diminishing returns.

**Sqft living** is the most influential feature. Predicted prices increase monotonically and substantially with living area, with the curve rising sharply for larger homes. This reflects the dominant role of size in real estate valuation. The rug marks reveal that the vast majority of observations fall in the lower-to-mid range; the right tail of the PDP, representing very large homes, is less data-supported.

**Floors** shows a positive but modest effect. Single-storey homes are associated with lower predicted prices, while two-storey homes command noticeably higher values. The relationship plateaus beyond 2 floors, indicating diminishing marginal returns. Because floors take only a few distinct values (1, 1.5, 2, 2.5, 3), the PDP produces a stepped pattern rather than a smooth curve.

---

# Conclusions

PDPs proved effective across all three exercises for extracting interpretable, globally-averaged feature effects from Random Forest models:

- In the **bike rental** task, all four features show directional effects consistent with real-world cycling behaviour — time driving growth, temperature and humidity shaping comfort, and wind discouraging use.
- The **2D PDP** confirms a reinforcing interaction between humidity and temperature, with warm and dry conditions producing peak predicted demand.
- In the **house price** task, living area dominates the prediction, while bedrooms, bathrooms and floors contribute secondary but interpretable signals aligned with market expectations.

A key limitation of PDPs is the **assumption of marginal independence**: when features are correlated, the grid evaluation integrates over combinations that may not exist in practice, potentially distorting the displayed effect. For datasets with strong feature correlations, Accumulated Local Effects (ALE) plots offer a more robust alternative.
